Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2017-02-16 manuscript/common_latex.tex
use the long names everywhere
add a second type III explanation for all genes
switch from 1:1 to single-copy ortholog
Include differential expression in non-humans
Include four additional tables in the supplemental material which
shows the differential expression of single-copy orthologs in each of
the species.

% species names
\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
\newcommand{\galili}[0]{\textit{S. galili}\xspace}
\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}

% top 50 gene table explanation. Note that if you modify this, you
% should probably also modify the human table explanation, which is
% slightly different.
\newcommand{\tableexplanation}[2]{The top 50 genes by FPKM in #1 which
  are not annotated as Mt\_rRNA, Mt\_tRNA, rRNA or whose \human
  ortholog is a housekeeping gene. Gene symbol is the gene symbol or
  Ensembl gene id. Human Symbol is the 1:1 human ortholog or the best
  human alignment when enclosed in parenthesis. Human Name is the HUGO
  name corresponding to the human symbol. A complete list of all genes
  is in \cref{#2}.}

@@

\newcommand{\lineagetableexplanation}[3]%
[A two-sided t test was used to calculate p values.]{%
  Genes whose expression changed in #2 as compared to #3. #1 FDR
  is the False Discovery Rate as estimated by the Benjamini and
  Hochberg procedure \cite{Benjamini1995:bh_procedure}.
  $\mathrm{log}_2$ FC is the base-2 logarithm of the fold change from
  #2 to #3 .}

\newcommand{\lineagetableexplanationtypeiii}[2]{%
  \lineagetableexplanation[A nested type-III anova was used to
  calculate p values, and the top 40 genes by $\text{log}_2$
  Fold-Change which changed significantly (FDR $\le 0.05$) are
  shown. ]{#1}{#2}}

>>>>

% species names
\newcommand{\smonkey}[0]{\textit{Ateles fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{Bos taurus}\xspace}
\newcommand{\dog}[0]{\textit{Canis familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{Dasypus novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{Equus caballus}\xspace}
\newcommand{\elephant}[0]{\textit{Loxodonta africana}\xspace}
\newcommand{\human}[0]{\textit{Homo sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{Monodelphis domestica}\xspace}
\newcommand{\mouse}[0]{\textit{Mus musculus}\xspace}
\newcommand{\galili}[0]{\textit{Spalax galili}\xspace}
\newcommand{\sheep}[0]{\textit{Ovis aries}\xspace}
\newcommand{\bonobo}[0]{\textit{Pan paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{Spalax carmeli}\xspace}
\newcommand{\pig}[0]{\textit{Sus scrofa}\xspace}

% top 50 gene table explanation. Note that if you modify this, you
% should probably also modify the human table explanation, which is
% slightly different.
\newcommand{\tableexplanation}[2]{The top 50 genes by FPKM in #1 which
  are not annotated as Mt\_rRNA, Mt\_tRNA, rRNA or whose \human
  ortholog is a housekeeping gene. Gene symbol is the gene symbol or
  Ensembl gene id. Human Symbol is the single-copy \human ortholog or the best
  human alignment when enclosed in parenthesis. Human Name is the HUGO
  name corresponding to the human symbol. A complete list of all genes
  is in \cref{#2}.}

@@

\newcommand{\lineagetableexplanation}[3]%
[A two-sided t test was used to calculate p values.]{%
  Genes whose expression changed in #2 as compared to #3. #1 FDR
  is the False Discovery Rate as estimated by the Benjamini and
  Hochberg procedure \cite{Benjamini1995:bh_procedure}.
  $\mathrm{log}_2$ FC is the base-2 logarithm of the fold change from
  #2 to #3.}

\newcommand{\lineagetableexplanationtypeiii}[2]{%
  \lineagetableexplanation[A nested type-III anova was used to
  calculate p values.]{#1}{#2}}

\newcommand{\lineagetableexplanationtypeiiishort}[2]{%
  \lineagetableexplanation[A nested type-III anova was used to
  calculate p values, and the top 40 genes by $\text{log}_2$
  Fold-Change which changed significantly (FDR $\le 0.05$) are
  shown. ]{#1}{#2}}

### 2017-02-16 manuscript/manuscript.Rnw
Include differential expression in non-humans
Include four additional tables in the supplemental material which
shows the differential expression of single-copy orthologs in each of
the species.

<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific","human_mouse_lineage_specific")))>>=
load("../data/placenta_lineage_specific")
load("../data/human_mouse_lineage_specific")
source("/home/don/projects/xtable/pkg/R/print.xtable.R")

@@

Using a type-III ANOVA to account for repeated testing of
\human and \mouse (see \cref{sec:comp-analys-gene} for details), we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}, Supplemental
\cref{tab:human_lineage_specific_all}). These genes include multiple
genes which are thought to be involved in pre-eclampsia. Notably,
\ac{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:crh_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\ac{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas, and is
involved in modulating trophoblast invasion~%
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}. Two other genes,
\ac{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \ac{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia~%
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

>>>>

<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(c(paste0("../data/",c("placenta_lineage_specific","human_mouse_lineage_specific")),"../literature_mining/human_lineage_vs_terms.txt"))>>=
load("../data/placenta_lineage_specific")
load("../data/human_mouse_lineage_specific")
source("/home/don/projects/xtable/pkg/R/print.xtable.R")

@@

Using a type-III ANOVA to account for repeated testing of
\human (see \cref{sec:comp-analys-gene} for details), we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}, Supplemental
\cref{tab:human_lineage_specific_all}). These genes include multiple
genes which are thought to be involved in pre-eclampsia. Notably,
\ac{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:crh_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\ac{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas, and is
involved in modulating trophoblast invasion~%
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}. Two other genes,
\ac{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \ac{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia~%
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

\subsection{Other Lineage and Clade Specific Genes}

We also compared differences in the \mouse lineage in comparison to
all other lineages using a type-III ANOVA as in the \human analysis,
and found few genes which are highly differentially expressed and are
also associated with adverse pregnancy outcomes. Comparisons between
rodentia (\mouse, \galili, and \carmeli) and all other species using
the mean of each species expression did not indicate any significant
differences (Supplemental \cref{tab:rodentia_lineage_specific}) nor
did the equivalent comparisons between laurasiatheria and all other
species (Supplemental \cref{tab:laurasiatheria_lineage_specific}) and
laurasiatheria and euarchontoglires (Supplemental
\cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
Additional placenta samples of each species and the collection of the
placentas of additional laurasiatherians may resolve the lack of
significance due to insufficient power.

### 2017-02-16 manuscript/supplemental_material.Rnw
Include differential expression in non-humans
Include four additional tables in the supplemental material which
shows the differential expression of single-copy orthologs in each of
the species.

placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=40,orderbyfc=FALSE,fdr.cutoff=1,caption=NULL,label=NULL) {
    temp.table <- 
        lineage.table[analysis.name==analysis &
                      p.value <= 0.05 &
                      fdr <= fdr.cutoff][order(p.value),
                                       list(human_name,p.value,fdr,fc),]
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
        if (orderbyfc) {
            display.table <- temp.table[order(-abs(`Fold Change`)),]
        }
        if (!is.na(members)) {
            members <- min(c(members,nrow(display.table)))
            display.table <- display.table[1:members,]
        }

        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        display.table[,`Human Name`:=NULL]
        # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
        print(xtable(display.table,
                     display=c("s","s","e","g","f"),
                     caption=caption,
                     label=label,
                     align=c("c","p{0.7in}","r","r","r")
                     ),
              booktabs=TRUE,
              floating=FALSE,
              caption.placement="top",
              tabular.environment=ifelse(is.na(members),"longtable","tabular"),
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              size="\\small",
              include.rownames=FALSE)

@@

<<human.lineage.specific,echo=FALSE,results="asis">>=
hum.all.lin.spec <-
    placenta.lineage.specific.table("hum.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    members=NA,
                                    caption=c("\\lineagetableexplanationtypeiii{\\human}{all other species}","\\human lineage specific genes"),
                                    label="tab:human_lineage_specific_all"
                                    )
setkey(hum.all.lin.spec,"Human Symbol")
@

>>>>

placenta.lineage.specific.table <- function(analysis,output=TRUE,
                                            lineage.table=placenta.lineage.specific.analysis,
                                            lit.mining=NULL,
                                            lit.mining.terms=NULL,
                                            members=40,
                                            orderbyfc=FALSE,
                                            fdr.cutoff=1,caption=NULL,
                                            label=NULL) {
    temp.table <- 
        lineage.table[analysis.name==analysis &
                      p.value <= 0.05 &
                      fdr <= fdr.cutoff][order(p.value),
                                       list(human_name,p.value,fdr,fc),]
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    if (!is.null(lit.mining)) {
        temp.table <-
            lit.mining[,list(human_name,`Adverse Pregnancy Outcomes`)
                       ][temp.table]
        temp.table[,
                   `Adv. Preg. Pub.`:="0"
                   ]
        temp.table[`Adverse Pregnancy Outcomes` >= 2,
               `Adv. Preg. Pub.`:="≥ 2"
               ]
        temp.table[`Adverse Pregnancy Outcomes` >= 5,
                   `Adv. Preg. Pub.`:="≥ 5"
                   ]
        temp.table[`Adverse Pregnancy Outcomes` >= 20,
                   `Adv. Preg. Pub.`:="≥ 20"
               ]
        temp.table[`Adverse Pregnancy Outcomes` == 1,
                   `Adv. Preg. Pub.`:="1"
                   ]
        temp.table[,`Adverse Pregnancy Outcomes`:=NULL]
    }
    keep.cols <-
        c("Approved Symbol","Approved Name","p.value","fdr","fc")
    if (!is.null(lit.mining)) {
        keep.cols <- c(keep.cols,"Adv. Preg. Pub.")
    }
    temp.table <- 
        hugo.nomenclature[temp.table][,keep.cols,with=FALSE]
    setnames(temp.table,
             c("Approved Symbol","Approved Name","p.value","fdr","fc"),
             c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
             )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
        col.disp <-
            c("s","s","e","g","f")
        col.align <-
            c("c",">{\\centering\\arraybackslash}p{1in}","r","r","r")
        if (!is.null(lit.mining)) {
            col.disp <- c(col.disp,"s")
            col.align <- c(col.align,">{\\centering\\arraybackslash}p{0.75in}")
        }
        if (orderbyfc) {
            display.table <- temp.table[order(-abs(`Fold Change`)),]
        }
        if (!is.na(members)) {
            members <- min(c(members,nrow(display.table)))
            display.table <- display.table[1:members,]
        }
        if (!is.null(lit.mining)) {
            display.table <-
                display.table[,`Adv. Preg. Pub.`:=
                                   paste0("\\href{https://www.ncbi.nlm.nih.gov/pubmed/?term=%22",
                                          `Human Symbol`,
                                          "%22[All%20Fields]%20AND%20",
                                          literature.mining.terms,"}{",
                                          `Adv. Preg. Pub.`,"}"
                                          )]
        }

        display.table <-
            display.table[,`Human Symbol`:=
                               paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",
                                      `Human Symbol`,"}{",
                                      "\\textit{",xtable.sanitize.text(`Human Symbol`),
                                      "}}")]
        display.table[,`Human Name`:=NULL]
                                        # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
        print(xtable(display.table,
                     display=col.disp,
                     caption=caption,
                     label=label,
                     align=col.align
                     ),
              booktabs=TRUE,
              floating=FALSE,
              caption.placement="top",
              tabular.environment=ifelse(is.na(members),"longtable","tabular"),
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              size="\\small",
              include.rownames=FALSE)

@@

\subsection{\human specific expression}

<<human.lineage.specific,echo=FALSE,results="asis",cache.extra=file.info("../literature_mining/human_lineage_vs_terms.txt")[,"mtime"]>>=
literature.mining <-
    data.table(read.table("../literature_mining/human_lineage_vs_terms.txt",
                          sep="\t",skip=0,stringsAsFactors=FALSE,
                          check.names=FALSE,header=TRUE),
               keep.rownames=TRUE)
setnames(literature.mining,"rn","human_name")
setkey(literature.mining,"human_name")
literature.mining[,`Adverse Pregnancy Outcomes`:=`preeclampsia`+
                       IUGR+`placenta previa`+`gestational diabetes`+
                       `premature birth`,
                  by=human_name]
literature.mining.terms <-
    paste0("%28%22preeclampsia%22[All%20Fields]%20OR%20",
           "%22pre-eclampsia%22[All%20Fields]%20OR%20",
           "%22eclampsia%22[All%20Fields]%20OR%20",
           "%22IUGR%22[All%20Fields]%20OR%20",
           "%22iugr%22[All%20Fields]%20OR%20",
           "%22intrauterine%20growth%20restriction%22[All%20Fields]%20OR%20",
           "%22placenta%20previa%22[All%20Fields]%20OR%20",
           "%22gestational%20diabetes%22[All%20Fields]%20OR%20",
           "%22premature%20birth%22[All%20Fields]%20OR%20",
           "%22premature%20labor%22[All%20Fields]%29")

hum.all.lin.spec <-
    placenta.lineage.specific.table("hum.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    lit.mining=literature.mining,
                                    lit.mining.terms=literature.mining.terms,
                                    members=NA,
                                    fdr.cutoff=0.05,
                                    orderbyfc=TRUE,
                                    caption=c("\\lineagetableexplanationtypeiii{\\human}{all other species}","\\human lineage specific genes"),
                                    label="tab:human_lineage_specific_all"
                                    )
setkey(hum.all.lin.spec,"Human Symbol")
@

\subsection{\mouse specific expression}

<<mouse.lineage.specific,echo=FALSE,results="asis",cache.extra=file.info("../literature_mining/mouse_lineage_vs_terms.txt")[,"mtime"]>>=
mouse.literature.mining <-
    data.table(read.table("../literature_mining/mouse_lineage_vs_terms.txt",
                          sep="\t",skip=0,stringsAsFactors=FALSE,
                          check.names=FALSE,header=TRUE),
               keep.rownames=TRUE)
setnames(mouse.literature.mining,"rn","human_name")
setkey(mouse.literature.mining,"human_name")
mouse.literature.mining[,`Adverse Pregnancy Outcomes`:=`preeclampsia`+
                             IUGR+`placenta previa`+`gestational diabetes`+
                             `premature birth`,
                        by=human_name]
mou.all.lin.spec <-
    placenta.lineage.specific.table("mou.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    lit.mining=mouse.literature.mining,
                                    lit.mining.terms=literature.mining.terms,
                                    fdr.cutoff=0.05,
                                    members=NA,
                                    orderbyfc=TRUE,
                                    caption=c("\\lineagetableexplanationtypeiii{\\mouse}{all other species}","\\mouse lineage specific genes"),
                                    label="tab:mouse_lineage_specific_all"
                                    )
setkey(mou.all.lin.spec,"Human Symbol")
@

\subsection{Rodentia, Laurasiatheria, and Euarchontoglires Comparisons}

\begin{table}[hb]
  \footnotesize
<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
rod.all.lin.spec <- placenta.lineage.specific.table("rod.all",fdr.cutoff=0.5)
setkey(rod.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanation{the rodentia order}{all other species}}
\label{tab:rodentia_lineage_specific}
\end{table}

\begin{table}[hb]
  \footnotesize
<<lau.lineage.specific,echo=FALSE,results="asis">>=
lau.all.lin.spec <- placenta.lineage.specific.table("lau.all",fdr.cutoff=0.5)
setkey(lau.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanation{the laurasiatheria
    superorder}{all other species}}
\label{tab:laurasiatheria_lineage_specific}
\end{table}

\begin{table}[hb]
  \footnotesize
<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau",fdr.cutoff=1)
setkey(eua.lau.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
\end{table}
